\\textsc{Mathematical Description}

### 2022-03-18 100-math-description/GR-sslcm-math-description.Rmd
Replace all \ref{} with \@ref()
This is general bookdown syntax and renders to word, HTML, and PDF

The model is framed in the Bayesian mode of inference, as it is intuitive for the necessary integration over the many layers of uncertainty (i.e., process vs. observational variability).
Further, Bayesian inference provides an intuitive framework to constrain population dynamics parameters to take on only biologically plausible values through the use of prior information -- we intentionally employ nearly exclusively uninformative prior information, but in several rare cases we were required to bring in auxiliary information in the form of reasonably strong priors (Table \ref{tab:param-table}).
A final benefit of using the Bayesian framework is the seamless propagation of uncertainty from directly estimated parameters to derived quantities (of which there are many) -- all unknown quantities, either estimated parameters or deterministic functions of estimated parameters, receive a posterior distribution in the Bayesian framework.
Due to the analytical intractability of the complexity of the model, we employ Markov Chain Monte Carlo methods (implemented via the JAGS probabilistic programming language) to sample from the joint posterior distribution of all unknown quantities in the model.

@@

We have described the life cycle model calculations as divided into three distinct life "phases" (Figure \ref{fig:diagram}): freshwater juvenile, ocean juvenile, and freshwater adult.
For each life phase, we describe the process model dynamics as a set of stochastic and deterministic processes followed by how they are informed by empirical data as part of the observation model.
Although we present these phases and model components in discrete sections, note that they are all components of a single integrated population dynamics model that is statistical in the sense that unknown quantities are estimated with uncertainty from input data.

@@

Process noise is introduced by assuming year-specific values are multivariate logit-normal random variables around the deterministic value.
This is done to enable modeling covariability in process noise among populations (or origins, in the few cases where population structure is not modeled).
The covariance matrix of these random variables are denoted by $\Sigma$, and have a common structure for all model components.
Take the hypothetical survival term $\phi^{A \rightarrow B}_{y,j}$ as an example.
We denote its random variability across years and populations as:
\begin{equation}
\logit\left(\boldsymbol{\phi}^{A \rightarrow B}_{y,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{A \rightarrow B}_{y,1:n_j}\right), \Sigma_{\phi^{A \rightarrow B}} \right]
\label{eq:generic-random}
\end{equation}
where $\Sigma_{\phi^{A \rightarrow B}}$ is an $n_j \times n_j$ matrix with the structure:
\begin{equation}
\left[
\begin{array}{cccc}
\left(\sigma_{\phi^{A \rightarrow B}_{j=1}}\right)^2 & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}} & \dots & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \left(\sigma_{\phi^{A \rightarrow B}_{j=2}}\right)^2 & \dots & \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\vdots & \vdots & \ddots & \vdots \\
\sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}} & \dots & \left(\sigma_{\phi^{A \rightarrow B}_{j=n_j}}\right)^2
\end{array}
\right]
\label{eq:generic-Sigma4by4}
\end{equation}
That is, all populations ($j$) have unique variance terms $(\sigma_{\phi^{A \rightarrow B}_j})^2$ but we assume the correlation component of the covariance calculation ($\rho_{\phi^{A \rightarrow B}}$) is identical for all population pairs.
In cases where a quantity varies annually and is different by origin but not by population, $\Sigma$ would have dimensions $n_o \times n_o$:
\begin{equation}
\left[
\begin{array}{cc}
\left(\sigma_{\phi^{A \rightarrow B}_{o=1}}\right)^2 & \sigma_{\phi^{A \rightarrow B}_{o=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{o=2}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\sigma_{\phi^{A \rightarrow B}_{o=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{o=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \left(\sigma_{\phi^{A \rightarrow B}_{o=2}}\right)^2
\end{array}
\right]
\label{eq:generic-Sigma2by2}
\end{equation}
Although the stochastic process requires the $\Sigma$ term, is is the component $\sigma_j$ (or $\sigma_o$) and $\rho$ terms that were assigned priors ($\mathcal{U}(0,5)$ and $\mathcal{U}(-1,1)$, respectively, Table \ref{tab:param-table}) and the $\Sigma$ matrices were constructed using the structures defined in eqs. \ref{eq:generic-Sigma4by4} and \ref{eq:generic-Sigma2by2}.

@@

Total egg production is the sum product of age-specific spawner abundance (from eq. \ref{eq:get-Sa}, below), proportion female, and fecundity:
\begin{equation}
E_{y,j} = \sum_o^{n_o} \sum_k^{n_k} S^{a}_{y,k,o,j} \cdot \Omega_{k,j} \cdot f_{k}
\label{eq:get-E}
\end{equation}
The proportion of spawners at age that are female ($\Omega_{k,j}$) was calculated from carcass survey data and averaged across years and fecundity ($f_k$) was calculated from samples taken from Grande Ronde-origin hatchery broodstock.
Both quantities were supplied to the model assuming no error or inter-annual variability.

@@

Realized (i.e., with process noise) egg-to-parr survival was a multivariate logit-normal random variable around the deterministic value ($\dot{\phi}^{E \rightarrow P^{b}}_{y,j}$) predicted by the Beverton-Holt function in eq. \ref{eq:get-phi-E-Pb-dot} with covariance matrix $\Sigma_{\phi^{E \rightarrow P^{b}}}$ (constructed of $\sigma_{\phi^{E \rightarrow P^{b}}_j}$ and $\rho_{\phi^{E \rightarrow P^{b}}}$ following eq. \ref{eq:generic-Sigma4by4}):
\begin{equation}
\logit\left(\boldsymbol{\phi}^{E \rightarrow P^{b}}_{y,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{E \rightarrow P^{b}}_{y,1:n_j}\right), \Sigma_{\phi^{E \rightarrow P^{b}}}\right]
\label{eq:phi-E-Pb-random}
\end{equation}
Parr recruitment ($P^{b}_{y,j}$) to the end of summer was then the product of total egg production and egg-to-parr survival:
\begin{equation}
P^{b}_{y,j} = \phi^{E \rightarrow P^{b}}_{y,j} \cdot E_{y,j}
\label{eq:get-Pb}
\end{equation}

@@

Not all summer parr ($P^{b}_{y,j}$) stay in the headwaters to rear overwinter -- some migrate out in the fall and rear farther downstream (termed "fall migrants" and indexed by $i = \mathrm{fall}$) and others migrate out the following spring (termed "spring migrants" and indexed by $i = \mathrm{spring}$).
Due to this difference in migratory phenology, these two groups of fish are monitored separately.
To enable building an observation model that could accommodate this feature of the populations in our model, summer parr was apportioned to these two strategies:
\begin{equation}
P^{a}_{y,i,j} = P^{b}_{y,j} \cdot \pi_{y,i,j}
\label{eq:get-Pa}
\end{equation}
where $P^{a}_{y,i,j}$ is migratory strategy-specific parr abundance and $\pi_{y,i,j}$ is the proportion that take on each migratory strategy.
For fall migrants, this proportion was a multivariate logit-normal random variable around a time-constant mean ($\dot{\pi}_{i = \mathrm{fall},j}$) with covariance matrix $\Sigma_{\pi_{i = \mathrm{fall}}}$ (constructed of $\sigma_{\pi_{i=\mathrm{fall},j}}$ and $\rho_{\pi_{i=\mathrm{fall}}}$ following eq. \ref{eq:generic-Sigma4by4}):
\begin{equation}
\logit\left(\boldsymbol{\pi}_{y,i = \mathrm{fall},1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\pi}}_{i = \mathrm{fall},1:n_j}\right), \Sigma_{\pi_{i = \mathrm{fall}}}\right]
\label{eq:pi-random}
\end{equation}
and the proportion that are spring migrants was obtained as its complement ($\pi_{y,i = \mathrm{spring},j} = 1 - \pi_{y,i = \mathrm{fall},j}$).

@@

Overwinter survival from the end of the parr year (i.e., age-1) to the start of the smolt year (i.e., age-2), denoted by $\phi^{P^{a} \rightarrow M^{b}}_{y,i,j}$, was modeled as a logit-linear function of parr density (scaled to the same metric of weighted usable habitat length [$\mathrm{WUL}_j$] used in eq. \ref{eq:beta-random}).
Deterministic overwinter survival was:
\begin{equation}
\logit\left(\dot{\phi}^{P^{a} \rightarrow M^{b}}_{y,i,j}\right) = \gamma_{0,i,j} + \gamma_{1,j} \cdot \left(\frac{P^{a}_{y,i,j}}{\mathrm{WUL}_j}\right)
\label{eq:get-phi-Pa-Mb-dot}
\end{equation}
Note the migratory strategy-specific intercept ($\gamma_{0,i,j}$) but common slope terms ($\gamma_{1,j}$).
Then, for each migratory strategy separately, overwinter survival was a multivariate logit-normal random variable around the deterministic logistic model prediction with covariance matrix $\Sigma_{\phi^{P^{a} \rightarrow M^{b}}_i}$ (each matrix constructed of $\sigma_{\phi^{P^{a} \rightarrow M^{b}}_{i,j}}$ and $\rho_{\phi^{P^{a} \rightarrow M^{b}}_{i}}$ following eq. \ref{eq:generic-Sigma4by4}):
\begin{equation}
\logit\left(\boldsymbol{\phi}^{P^{a} \rightarrow M^{b}}_{y,i,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{P^{a} \rightarrow M^{b}}_{y,i,1:n_j}\right), \Sigma_{\phi^{P^{a} \rightarrow M^{b}}_i}\right]
\label{eq:phi-Pa-Mb-random}
\end{equation}
Natural-origin (indexed by $o=\mathrm{NOR}$) parr surviving to the beginning of the following year was then:
\begin{equation}
M^{b}_{y,i,o=\mathrm{NOR},j} = P^{a}_{y,i,j} \cdot \phi^{P^{a} \rightarrow M^{b}}_{y,i,j}
\label{eq:get-Mb-NOR}
\end{equation}
For populations with hatchery supplementation, hatchery-origin smolt releases (assumed known without error) were introduced at this point as well and stored in the variable $M^{b}_{y,i=\mathrm{spring},o=\mathrm{HOR},j}$; we assumed no hatchery-origin parr were released the prior year, thus $M^{b}_{y,i=\mathrm{fall},o=\mathrm{HOR},j} = 0$.

@@

In-basin smolt ($M^{b}_{y,i,o,j}$) migrate to the ocean in the spring of the year they become age-2.
We divide mortality experienced during this migration into two stages: (_i_) from in-basin to Lower Granite Dam (LGR), the first in a series of eight dams encountered in the downstream migration, and (_ii_) from LGR to the ocean.
For parameter identifiability in mortality stage (_i_), we were required to assume that survival from in-basin to LGR was identical between natural-origin fall and spring migrants (i.e., $\phi^{M^{b} \rightarrow M^{a}}_{y,i=\mathrm{fall},o=\mathrm{NOR},j} = \phi^{M^{b} \rightarrow M^{a}}_{y,i=\mathrm{spring},o=\mathrm{NOR},j}$, we thus omit the $i$ dimension when defining these parameter below), but we were able to model natural-origin survival separately from hatchery-origin.
We assume that year-specific realized survival values were multivariate logit-normal random variables around a time-constant and density-independent deterministic value ($\dot{\phi}^{M^{b} \rightarrow M^{a}}_{o,j}$) with covariance matrix $\Sigma_{\phi^{M^{b} \rightarrow M^{a}}_o}$ (constructed of $\sigma_{\phi^{M^{b} \rightarrow M^{a}}_{o,j}}$ and $\rho_{\phi^{M^{b} \rightarrow M^{a}}_{o}}$ following eq. \ref{eq:generic-Sigma4by4}):
\begin{equation}
\logit\left(\boldsymbol{\phi}^{M^{b} \rightarrow M^{a}}_{y,o,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{M^{b} \rightarrow M^{a}}_{o,1:n_j}\right), \Sigma_{\phi^{M^{b} \rightarrow M^{a}}_o}\right]
\label{eq:phi-Mb-Ma-random}
\end{equation}
Thus, the abundance of smolt reaching LGR was:
\begin{equation}
M^{a}_{y,i,o,j} = M^{b}_{y,i,o,j} \cdot \phi^{M^{b} \rightarrow M^{a}}_{y,o,j}
\label{eq:get-Ma}
\end{equation}
After reaching LGR, we modeled survival during the downstream migration through the hydrosystem on the Snake and Columbia rivers as a single survival rate, specific to each year and origin but shared among populations ($\phi^{M^{a} \rightarrow O^{0}}_{y,o}$).
We assumed these survival rates were multivariate logit-normal random variables around a time-constant and density-independent deterministic value ($\dot{\phi}^{M^{a} \rightarrow O^{0}}_{o}$) with covariance matrix $\Sigma_{\phi^{M^{a} \rightarrow O^{0}}}$ (constructed of $\sigma_{\phi^{M^{a} \rightarrow O^{0}}_{o}}$ and $\rho_{\phi^{M^{a} \rightarrow O^{0}}}$ following eq. \ref{eq:generic-Sigma2by2}):
\begin{equation}
\logit\left(\boldsymbol{\phi}^{M^{a} \rightarrow O^{0}}_{y,1:n_o}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{M^{a} \rightarrow O^{0}}_{1:n_o}\right), \Sigma_{\phi^{M^{a} \rightarrow O^{0}}}\right]
\label{eq:phi-Ma-O0-random}
\end{equation}
Thus, the abundance of fish reaching the ocean by origin was:
\begin{equation}
O^{0}_{y,o,j} = \phi^{M^{a} \rightarrow O^{0}}_{y,o} \sum_i^{n_i} M^{a}_{y,i,o,j}
\label{eq:get-O0}
\end{equation}
Note the summation across migratory strategies ($i$) in eq. \ref{eq:get-O0}, which indicates that the fates of fish with these different strategies were assumed to be the result of identical processes following arrival to LGR.

@@

Now, we must place the maturing portion of each of $O^{1}_{y,o,j}$, $O^{2}_{y,o,j}$, and $O^{3}_{y,o,j}$ into the correct year and age of return to the mouth of the Columbia River ($R^{b}_{y,k,o,j}$).
Fish spawned in year $y$ will return as the $k^{\mathrm{th}}$ possible age at maturity in year $y + k_{\mathrm{min}} + k - 1$ (age-3 is the first possible total age of maturity, thus $k_{\mathrm{min}}$ = 3).
As an example, fish spawned in brood year $y$ = 2000 will return in 2003 as age-3 ($k$ = 1), in 2004 as age-4 ($k$ = 2), or in 2005 as age-5 ($k$ = 3).
Thus, the abundance of fish spawned in brood year $y$ returning as age-3, age-4, and age-5, respectively, was:
\begin{equation}
\begin{aligned}
R^{b}_{y+k_{\mathrm{min}} + 1 - 1,1,o,j} &= O^{1}_{y,o,j} \cdot \psi^{O^{1}}_{y,o,j} \\
R^{b}_{y+k_{\mathrm{min}} + 2 - 1,2,o,j} &= O^{2}_{y,o,j} \cdot \psi^{O^{2}}_{y,o,j} \\
R^{b}_{y+k_{\mathrm{min}} + 3 - 1,3,o,j} &= O^{3}_{y,o,j} \cdot \psi^{O^{3}}_{y,o,j} \\
\end{aligned}
\label{eq:get-Rbs}
\end{equation}
Note, however, that this leaves 12 age/year combinations per origin unpopulated (the first 3 missing years for age-3 returns, first 4 missing years for age-4 returns, and first 5 missing years for age-5 returns).
This is because the life cycle dynamics must start somewhere, and no juvenile process model outcomes existed that would ultimately become these $R^{b}_{y,k,o,j}$ values in these early years.
Thus, to initialize the adult returns for natural-origin fish (and as a result, the rest of the life cycle), we estimated these 12 year/age return abundances as free parameters with fairly restrictive priors (Table \ref{tab:param-table}) with boundaries loosely informed by the ranges of adult returns at age observed in the early years of the data time series.
Initial abundance of hatchery-origin returns was handled by the "straying model", described below.

>>>>

The model is framed in the Bayesian mode of inference, as it is intuitive for the necessary integration over the many layers of uncertainty (i.e., process vs. observational variability).
Further, Bayesian inference provides an intuitive framework to constrain population dynamics parameters to take on only biologically plausible values through the use of prior information -- we intentionally employ nearly exclusively uninformative prior information, but in several rare cases we were required to bring in auxiliary information in the form of reasonably strong priors (Table \@ref(tab:param-table)).
A final benefit of using the Bayesian framework is the seamless propagation of uncertainty from directly estimated parameters to derived quantities (of which there are many) -- all unknown quantities, either estimated parameters or deterministic functions of estimated parameters, receive a posterior distribution in the Bayesian framework.
Due to the analytical intractability of the complexity of the model, we employ Markov Chain Monte Carlo methods (implemented via the JAGS probabilistic programming language) to sample from the joint posterior distribution of all unknown quantities in the model.

@@

We have described the life cycle model calculations as divided into three distinct life "phases" (Figure \@ref(fig:diagram)): freshwater juvenile, ocean juvenile, and freshwater adult.
For each life phase, we describe the process model dynamics as a set of stochastic and deterministic processes followed by how they are informed by empirical data as part of the observation model.
Although we present these phases and model components in discrete sections, note that they are all components of a single integrated population dynamics model that is statistical in the sense that unknown quantities are estimated with uncertainty from input data.

@@

Process noise is introduced by assuming year-specific values are multivariate logit-normal random variables around the deterministic value.
This is done to enable modeling covariability in process noise among populations (or origins, in the few cases where population structure is not modeled).
The covariance matrix of these random variables are denoted by $\Sigma$, and have a common structure for all model components.
Take the hypothetical survival term $\phi^{A \rightarrow B}_{y,j}$ as an example.
We denote its random variability across years and populations as:
\begin{equation}
\logit\left(\boldsymbol{\phi}^{A \rightarrow B}_{y,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{A \rightarrow B}_{y,1:n_j}\right), \Sigma_{\phi^{A \rightarrow B}} \right]
\label{eq:generic-random}
\end{equation}
where $\Sigma_{\phi^{A \rightarrow B}}$ is an $n_j \times n_j$ matrix with the structure:
\begin{equation}
\left[
\begin{array}{cccc}
\left(\sigma_{\phi^{A \rightarrow B}_{j=1}}\right)^2 & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}} & \dots & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \left(\sigma_{\phi^{A \rightarrow B}_{j=2}}\right)^2 & \dots & \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\vdots & \vdots & \ddots & \vdots \\
\sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}} & \dots & \left(\sigma_{\phi^{A \rightarrow B}_{j=n_j}}\right)^2
\end{array}
\right]
\label{eq:generic-Sigma4by4}
\end{equation}
That is, all populations ($j$) have unique variance terms $(\sigma_{\phi^{A \rightarrow B}_j})^2$ but we assume the correlation component of the covariance calculation ($\rho_{\phi^{A \rightarrow B}}$) is identical for all population pairs.
In cases where a quantity varies annually and is different by origin but not by population, $\Sigma$ would have dimensions $n_o \times n_o$:
\begin{equation}
\left[
\begin{array}{cc}
\left(\sigma_{\phi^{A \rightarrow B}_{o=1}}\right)^2 & \sigma_{\phi^{A \rightarrow B}_{o=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{o=2}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\sigma_{\phi^{A \rightarrow B}_{o=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{o=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \left(\sigma_{\phi^{A \rightarrow B}_{o=2}}\right)^2
\end{array}
\right]
\label{eq:generic-Sigma2by2}
\end{equation}
Although the stochastic process requires the $\Sigma$ term, is is the component $\sigma_j$ (or $\sigma_o$) and $\rho$ terms that were assigned priors ($\mathcal{U}(0,5)$ and $\mathcal{U}(-1,1)$, respectively, Table \ref{tab:param-table}) and the $\Sigma$ matrices were constructed using the structures defined in eqs. \ref{eq:generic-Sigma4by4} and \ref{eq:generic-Sigma2by2}.
Although the stochastic process requires the $\Sigma$ term, is is the component $\sigma_j$ (or $\sigma_o$) and $\rho$ terms that were assigned priors ($\mathcal{U}(0,5)$ and $\mathcal{U}(-1,1)$, respectively, Table \@ref(tab:param-table)) and the $\Sigma$ matrices were constructed using the structures defined in eqs. \@ref(eq:generic-Sigma4by4) and \@ref(eq:generic-Sigma2by2).

@@

Total egg production is the sum product of age-specific spawner abundance (from eq. \@ref(eq:get-Sa), below), proportion female, and fecundity:
\begin{equation}
E_{y,j} = \sum_o^{n_o} \sum_k^{n_k} S^{a}_{y,k,o,j} \cdot \Omega_{k,j} \cdot f_{k}
\label{eq:get-E}
\end{equation}
The proportion of spawners at age that are female ($\Omega_{k,j}$) was calculated from carcass survey data and averaged across years and fecundity ($f_k$) was calculated from samples taken from Grande Ronde-origin hatchery broodstock.
Both quantities were supplied to the model assuming no error or inter-annual variability.

@@

Realized (i.e., with process noise) egg-to-parr survival was a multivariate logit-normal random variable around the deterministic value ($\dot{\phi}^{E \rightarrow P^{b}}_{y,j}$) predicted by the Beverton-Holt function in eq. \@ref(eq:get-phi-E-Pb-dot) with covariance matrix $\Sigma_{\phi^{E \rightarrow P^{b}}}$ (constructed of $\sigma_{\phi^{E \rightarrow P^{b}}_j}$ and $\rho_{\phi^{E \rightarrow P^{b}}}$ following eq. \@ref(eq:generic-Sigma4by4)):
\begin{equation}
\logit\left(\boldsymbol{\phi}^{E \rightarrow P^{b}}_{y,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{E \rightarrow P^{b}}_{y,1:n_j}\right), \Sigma_{\phi^{E \rightarrow P^{b}}}\right]
\label{eq:phi-E-Pb-random}
\end{equation}
Parr recruitment ($P^{b}_{y,j}$) to the end of summer was then the product of total egg production and egg-to-parr survival:
\begin{equation}
P^{b}_{y,j} = \phi^{E \rightarrow P^{b}}_{y,j} \cdot E_{y,j}
\label{eq:get-Pb}
\end{equation}

@@

Not all summer parr ($P^{b}_{y,j}$) stay in the headwaters to rear overwinter -- some migrate out in the fall and rear farther downstream (termed "fall migrants" and indexed by $i = \mathrm{fall}$) and others migrate out the following spring (termed "spring migrants" and indexed by $i = \mathrm{spring}$).
Due to this difference in migratory phenology, these two groups of fish are monitored separately.
To enable building an observation model that could accommodate this feature of the populations in our model, summer parr was apportioned to these two strategies:
\begin{equation}
P^{a}_{y,i,j} = P^{b}_{y,j} \cdot \pi_{y,i,j}
\label{eq:get-Pa}
\end{equation}
where $P^{a}_{y,i,j}$ is migratory strategy-specific parr abundance and $\pi_{y,i,j}$ is the proportion that take on each migratory strategy.
For fall migrants, this proportion was a multivariate logit-normal random variable around a time-constant mean ($\dot{\pi}_{i = \mathrm{fall},j}$) with covariance matrix $\Sigma_{\pi_{i = \mathrm{fall}}}$ (constructed of $\sigma_{\pi_{i=\mathrm{fall},j}}$ and $\rho_{\pi_{i=\mathrm{fall}}}$ following eq. \@ref(eq:generic-Sigma4by4)):
\begin{equation}
\logit\left(\boldsymbol{\pi}_{y,i = \mathrm{fall},1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\pi}}_{i = \mathrm{fall},1:n_j}\right), \Sigma_{\pi_{i = \mathrm{fall}}}\right]
\label{eq:pi-random}
\end{equation}
and the proportion that are spring migrants was obtained as its complement ($\pi_{y,i = \mathrm{spring},j} = 1 - \pi_{y,i = \mathrm{fall},j}$).

@@

Overwinter survival from the end of the parr year (i.e., age-1) to the start of the smolt year (i.e., age-2), denoted by $\phi^{P^{a} \rightarrow M^{b}}_{y,i,j}$, was modeled as a logit-linear function of parr density (scaled to the same metric of weighted usable habitat length [$\mathrm{WUL}_j$] used in eq. \@ref(eq:beta-random)).
Deterministic overwinter survival was:
\begin{equation}
\logit\left(\dot{\phi}^{P^{a} \rightarrow M^{b}}_{y,i,j}\right) = \gamma_{0,i,j} + \gamma_{1,j} \cdot \left(\frac{P^{a}_{y,i,j}}{\mathrm{WUL}_j}\right)
\label{eq:get-phi-Pa-Mb-dot}
\end{equation}
Note the migratory strategy-specific intercept ($\gamma_{0,i,j}$) but common slope terms ($\gamma_{1,j}$).
Then, for each migratory strategy separately, overwinter survival was a multivariate logit-normal random variable around the deterministic logistic model prediction with covariance matrix $\Sigma_{\phi^{P^{a} \rightarrow M^{b}}_i}$ (each matrix constructed of $\sigma_{\phi^{P^{a} \rightarrow M^{b}}_{i,j}}$ and $\rho_{\phi^{P^{a} \rightarrow M^{b}}_{i}}$ following eq. \@ref(eq:generic-Sigma4by4)):
\begin{equation}
\logit\left(\boldsymbol{\phi}^{P^{a} \rightarrow M^{b}}_{y,i,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{P^{a} \rightarrow M^{b}}_{y,i,1:n_j}\right), \Sigma_{\phi^{P^{a} \rightarrow M^{b}}_i}\right]
\label{eq:phi-Pa-Mb-random}
\end{equation}
Natural-origin (indexed by $o=\mathrm{NOR}$) parr surviving to the beginning of the following year was then:
\begin{equation}
M^{b}_{y,i,o=\mathrm{NOR},j} = P^{a}_{y,i,j} \cdot \phi^{P^{a} \rightarrow M^{b}}_{y,i,j}
\label{eq:get-Mb-NOR}
\end{equation}
For populations with hatchery supplementation, hatchery-origin smolt releases (assumed known without error) were introduced at this point as well and stored in the variable $M^{b}_{y,i=\mathrm{spring},o=\mathrm{HOR},j}$; we assumed no hatchery-origin parr were released the prior year, thus $M^{b}_{y,i=\mathrm{fall},o=\mathrm{HOR},j} = 0$.

@@

In-basin smolt ($M^{b}_{y,i,o,j}$) migrate to the ocean in the spring of the year they become age-2.
We divide mortality experienced during this migration into two stages: (_i_) from in-basin to Lower Granite Dam (LGR), the first in a series of eight dams encountered in the downstream migration, and (_ii_) from LGR to the ocean.
For parameter identifiability in mortality stage (_i_), we were required to assume that survival from in-basin to LGR was identical between natural-origin fall and spring migrants (i.e., $\phi^{M^{b} \rightarrow M^{a}}_{y,i=\mathrm{fall},o=\mathrm{NOR},j} = \phi^{M^{b} \rightarrow M^{a}}_{y,i=\mathrm{spring},o=\mathrm{NOR},j}$, we thus omit the $i$ dimension when defining these parameter below), but we were able to model natural-origin survival separately from hatchery-origin.
We assume that year-specific realized survival values were multivariate logit-normal random variables around a time-constant and density-independent deterministic value ($\dot{\phi}^{M^{b} \rightarrow M^{a}}_{o,j}$) with covariance matrix $\Sigma_{\phi^{M^{b} \rightarrow M^{a}}_o}$ (constructed of $\sigma_{\phi^{M^{b} \rightarrow M^{a}}_{o,j}}$ and $\rho_{\phi^{M^{b} \rightarrow M^{a}}_{o}}$ following eq. \@ref(eq:generic-Sigma4by4)):
\begin{equation}
\logit\left(\boldsymbol{\phi}^{M^{b} \rightarrow M^{a}}_{y,o,1:n_j}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{M^{b} \rightarrow M^{a}}_{o,1:n_j}\right), \Sigma_{\phi^{M^{b} \rightarrow M^{a}}_o}\right]
\label{eq:phi-Mb-Ma-random}
\end{equation}
Thus, the abundance of smolt reaching LGR was:
\begin{equation}
M^{a}_{y,i,o,j} = M^{b}_{y,i,o,j} \cdot \phi^{M^{b} \rightarrow M^{a}}_{y,o,j}
\label{eq:get-Ma}
\end{equation}
After reaching LGR, we modeled survival during the downstream migration through the hydrosystem on the Snake and Columbia rivers as a single survival rate, specific to each year and origin but shared among populations ($\phi^{M^{a} \rightarrow O^{0}}_{y,o}$).
We assumed these survival rates were multivariate logit-normal random variables around a time-constant and density-independent deterministic value ($\dot{\phi}^{M^{a} \rightarrow O^{0}}_{o}$) with covariance matrix $\Sigma_{\phi^{M^{a} \rightarrow O^{0}}}$ (constructed of $\sigma_{\phi^{M^{a} \rightarrow O^{0}}_{o}}$ and $\rho_{\phi^{M^{a} \rightarrow O^{0}}}$ following eq. \@ref(eq:generic-Sigma2by2)):
\begin{equation}
\logit\left(\boldsymbol{\phi}^{M^{a} \rightarrow O^{0}}_{y,1:n_o}\right) \sim \mathcal{MVN}\left[\logit\left(\boldsymbol{\dot{\phi}}^{M^{a} \rightarrow O^{0}}_{1:n_o}\right), \Sigma_{\phi^{M^{a} \rightarrow O^{0}}}\right]
\label{eq:phi-Ma-O0-random}
\end{equation}
Thus, the abundance of fish reaching the ocean by origin was:
\begin{equation}
O^{0}_{y,o,j} = \phi^{M^{a} \rightarrow O^{0}}_{y,o} \sum_i^{n_i} M^{a}_{y,i,o,j}
\label{eq:get-O0}
\end{equation}
Note the summation across migratory strategies ($i$) in eq. \@ref(eq:get-O0), which indicates that the fates of fish with these different strategies were assumed to be the result of identical processes following arrival to LGR.

@@

Now, we must place the maturing portion of each of $O^{1}_{y,o,j}$, $O^{2}_{y,o,j}$, and $O^{3}_{y,o,j}$ into the correct year and age of return to the mouth of the Columbia River ($R^{b}_{y,k,o,j}$).
Fish spawned in year $y$ will return as the $k^{\mathrm{th}}$ possible age at maturity in year $y + k_{\mathrm{min}} + k - 1$ (age-3 is the first possible total age of maturity, thus $k_{\mathrm{min}}$ = 3).
As an example, fish spawned in brood year $y$ = 2000 will return in 2003 as age-3 ($k$ = 1), in 2004 as age-4 ($k$ = 2), or in 2005 as age-5 ($k$ = 3).
Thus, the abundance of fish spawned in brood year $y$ returning as age-3, age-4, and age-5, respectively, was:
\begin{equation}
\begin{aligned}
R^{b}_{y+k_{\mathrm{min}} + 1 - 1,1,o,j} &= O^{1}_{y,o,j} \cdot \psi^{O^{1}}_{y,o,j} \\
R^{b}_{y+k_{\mathrm{min}} + 2 - 1,2,o,j} &= O^{2}_{y,o,j} \cdot \psi^{O^{2}}_{y,o,j} \\
R^{b}_{y+k_{\mathrm{min}} + 3 - 1,3,o,j} &= O^{3}_{y,o,j} \cdot \psi^{O^{3}}_{y,o,j} \\
\end{aligned}
\label{eq:get-Rbs}
\end{equation}
Note, however, that this leaves 12 age/year combinations per origin unpopulated (the first 3 missing years for age-3 returns, first 4 missing years for age-4 returns, and first 5 missing years for age-5 returns).
This is because the life cycle dynamics must start somewhere, and no juvenile process model outcomes existed that would ultimately become these $R^{b}_{y,k,o,j}$ values in these early years.
Thus, to initialize the adult returns for natural-origin fish (and as a result, the rest of the life cycle), we estimated these 12 year/age return abundances as free parameters with fairly restrictive priors (Table \@ref(tab:param-table)) with boundaries loosely informed by the ranges of adult returns at age observed in the early years of the data time series.
Initial abundance of hatchery-origin returns was handled by the "straying model", described below.
